Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-05-24 src/tf-cutcell-comparison/fig-gravityWaves-theta.tex
move into src directory

\includegraphics[width=2.2in]{../gravityWaves-snapCol-250m/18000/thetaDiff.pdf}
\hspace{0.1in}
\includegraphics[width=4in]{../gravityWaves-snapCol-250m/18000/thetaDiffZoom.pdf}
\vspace*{0.1in}

>>>>

\includegraphics[width=2.2in]{../gravityWaves-cutCell-250m/18000/thetaDiff.pdf}
\hspace{0.1in}
\includegraphics[width=4in]{../gravityWaves-cutCell-250m/18000/thetaDiffZoom.pdf}
\vspace*{0.1in}

### 2015-05-24 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
move into src directory

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation.  Orography also exerts strong non-local influences by the atmospheric response to form drag and gravity wave drag.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.  Because most models are designed for a particular type of grid, existing studies of cut cell solutions have compared results with terrain following grid solutions implemented within different models, for example \citet{good2014}.  Instead, this study uses a single model to enable a like-for-like comparison between solutions using terrain following and cut cell grids.

@@

The OpenFOAM utility \texttt{snappyHexMesh} was used to create a grid that approximates the cut cell method.  First, a custom utility was used to move points beneath the surface up to the surface creating small cells near mountain peaks.  The utility is available at \url{https://github.com/hertzsprung/AtmosFOAM/tree/2386ffb/applications/utilities/mesh/add2dMountain}.  \TODO{could just link to the whole of ExnerFoam/AtmosTests} Second, the surface faces were taken from the BTF grid and \texttt{snappyHexMesh} was used to intersect the surface with the grid.  This tool removes cells whose centres are below the surface and displaces boundary vertices so that they are `snapped' to the BTF surface \citep{openfoam2015}.  An example of the resulting grid is shown in figure~\ref{fig:grid}c.

@@

The model has a curl-free pressure gradient formulation and an upwind-biased cubic advection scheme.  It uses a Lorenz C-grid staggering of thermodynamic variables such that $\theta$, $\rho$ and $\Pi$ are stored at cell centres and velocities at cell faces.

@@

@@

By \(t = \SI{10000}{\second}\), the tracer suffers from distortion on the BTF grid using the centred linear scheme and some artefacts remain about the mountain peak.  The tracer has spread vertically due to increased numerical errors when the tracer is transported between layers.  Distortions are reduced by using the fourth order scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}c), but the computational mode is seen as a grid-scale oscillation that travels in the opposite direction to the wind (figure~\ref{fig:advection-tracer}e).  The results from the centred linear and fourth order schemes are slightly worse than the respective results from \citet{schaer2002} (their figure 6a and figure 8) because gradients in $\Psi$ have not been removed by interpolation.

@@

The results of the tracer advection test show that numerical errors are due primary to misalignment of the flow with the grid.  Using the upwind-biased cubic scheme, distortions in the grid do not significantly distort the tracer.

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-advection-tracer/fig-advection-tracer.pdf}
%
	\caption{Horizontally advected tracer contours at \(t = \SI{0}{\second}\), \SI{5000}{\second} and \SI{10000}{\second} using (a) centred linear scheme on a BTF grid, the upwind-biased cubic scheme on (b) the cut cell grid, (d) the BTF grid, and (c) the fourth order scheme from \citet{schaer2002} on the BTF grid, with contour intervals every 0.1.  Errors on the BTF grid at \(t = \SI{10000}{\second}\) are shown for (e) the fourth order scheme from \citet{schaer2002}, and (f) the upwind-biased cubic scheme, with contour intervals every 0.01.  Negative contours denoted by dotted lines.  The terrain profile is also shown immediately above the $x$ axis.}
	\label{fig:advection-tracer}
\end{figure*}

\subsection{Terrain following advection}
In the horizontal advection test, results were more accurate where the flow was aligned with the grid layers, and distortions in the BTF grid led to increased errors.  We formulate a new tracer advection test in which the velocity field is everywhere tangential to the basic terrain following coordinate surfaces.  In this new test, flow is aligned with the BTF grid layers.  Misalignment on the SLEVE and cut cell grids is designed to challenge the advection scheme.

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes and, for the horizontal advection test, the fourth order scheme from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{l l l l S S S S}
\hline\hline
                  &                     &                  & Analytic & {BTF}   & {SLEVE}         & {Cut cell}     & {No terrain} \\
\hline
Horizontal        & Centred linear      & \(\ell^2\) error & 0        & 0.0210  & 0.00233         & 0.00224        & 0.00223     \\
                  &                     & min              & 0        & -0.275  & -0.0252         & -0.0251        & -0.0251     \\
                  &                     & max              & 1        & 0.925   & 0.985           & 0.985          & 0.985       \\
                  & Fourth order        & min              & 0        & -0.0926 & -0.00174        & {---}          & -0.00178    \\
                  &                     & max              & 1        & 1.00    & 0.984           & {---}          & 0.983       \\
                  & Cubic upwind-biased & \(\ell^2\) error & 0        & 0.00791 & 0.00108         & 0.000577       & 0.000576    \\
                  &                     & min              & 0        & -0.0446 & -0.0106         & -0.000674      & -0.00674    \\
                  &                     & max              & 1        & 0.925   & 0.982           & 0.983          & 0.983       \\
\hline
Terrain following & Centred linear      & \(\ell^2\) error & 0        & 0.00251 & 0.0173          & 13.6           & {---}        \\
		  &                     & min              & 0        & -0.0245 & -0.120          & -623           & {---}        \\
                  &                     & max              & 1        & 0.985   & 0.950           & 3480           & {---}        \\
		  & Cubic upwind-biased & \(\ell^2\) error & 0        & 0.00154 & 0.0120          & 0.0134         & {---}        \\
                  &                     & min              & 0        & -0.0110 & -0.0263         & -0.028         & {---}        \\
                  &                     & max              & 1        & 0.983   & 0.865           & 0.851          & {---}        \\
\hline
\end{tabular}
\end{table*}

@@

The test was integrated forward by 5 hours on the BTF, SLEVE and cut cell grids, and a regular grid with flat terrain.  Maximum vertical velocities are presented in figure~\ref{fig:resting}.  In agreement with \citet{klemp2011}, vertical velocities are larger on more distorted grids.  However, magnitudes are smaller comparing results on the terrain following grids with those from \citet{klemp2011}.  Using the model from \citet{weller-shahrokhi2014}, which includes a curl-free pressure gradient formulation, $w$ reaches a maximum of \(\sim \SI{0.35}{\meter\per\second}\) on the BTF grid, compared with a maximum of \(\sim \SI{7}{\meter\per\second}\) found by \citet{klemp2011} using their improved horizontal pressure gradient formulation.

Unlike the result from \citet{klemp2011}, the SLEVE grid does not significantly reduce vertical velocities compared to the BTF grid.  However, errors are three orders of magnitude smaller on the cut cell grid with vertical velocities of \(\sim \SI{1e-3}{\meter\per\second}\).  The smallest error of \(\sim \SI{1e-10}{\meter\per\second}\) is found on the regular grid.

@@

Test results are compared between the BTF, SLEVE and cut cell grids.  Vertical velocities on the BTF grid are shown in figure~\ref{fig:gw-w}a, which have few visible differences from results on the SLEVE and cut cell grids (not shown).  Vertical velocities on all grids are in agreement with the high resolution solution from \citet{melvin2010}, shown in figure~\ref{fig:gw-w}b.

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves/fig-gravityWaves.pdf}
%
	\caption{Vertical cross section of vertical velocity contours in the gravity waves test after 5 hours with mountain height $h_0 = \SI{250}{\meter}$ on (a) the BTF grid compared with (b) the mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.  Contours are every \SI{5e-2}{\meter\per\second} with solid lines denoting ascent and dashed lines descent.  Small oscillations in contour lines are due to contouring on an irregular grid.}
	\label{fig:gw-w}
\end{figure}

>>>>

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation.  Orography also exerts strong non-local influences from the heat release due to convection and by the atmospheric response to form drag and gravity wave drag.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.  Because most models are designed for a particular type of grid, existing studies of cut cell solutions have compared results with terrain following grid solutions implemented within different models, for example \citet{good2014}.  Instead, this study uses a single model to enable a like-for-like comparison between solutions using terrain following and cut cell grids.

@@

The OpenFOAM utility \texttt{snappyHexMesh} was used to create a grid that approximates the cut cell method.  First, a custom utility was used to move points beneath the surface up to the surface creating small cells near mountain peaks.  Second, the surface faces were taken from the BTF grid and \texttt{snappyHexMesh} was used to intersect the surface with the grid.  This tool removes cells whose centres are below the surface and displaces boundary vertices so that they are `snapped' to the BTF surface \citep{openfoam2015}.  An example of the resulting grid is shown in figure~\ref{fig:grid}c.

@@

The model uses a Lorenz C-grid staggering of thermodynamic variables such that $\theta$, $\rho$ and $\Pi$ are stored at cell centres and velocities at cell faces.

A minor change has been made to the curl-free, non-orthogonal formulation as described in \citet{weller-shahrokhi2014} to improve stability on non-orthogonal grids.  \citet{weller-shahrokhi2014} use a Hodge dual operator, \(H\), to map from co-variant to contra-variant velocity components:
\begin{align}
	U = HV
\end{align}
where \(U = \rho\bm{u}\cdot\bm{S}_f\) and \(V=\rho\bm{u}\cdot\bm{d}_f\).  \(\bm{S}_f\) is the surface normal vector for a face, \(f\), which has a magnitude of the surface area, and \(\bm{d}_f\) is the vector between neighbouring cell centres.

In order to create a semi-implicit formulation, \(H\) is split into diagonal and off-diagonal components and the off-diagonal components are treated explicitly in the Helmholtz equation.  However, this was found to be unstable when using a large timestep on a grid with large non-orthogonality.  Improved stability is achieved by splitting \(H\) into the orthogonal part and the non-orthogonal correction: 
\begin{align}
	H = H_c + H_\mathrm{corr}
\end{align}
where the diagonal matrix \(H_c=\frac{|\bm{S}_f|}{|\bm{d}_f|}\) and the non-orthogonal correction is simply \(H_\mathrm{corr} = H - H_c\).  The orthogonal part, \(H_c\), is treated implicitly in the Helmholtz equation and the non-orthogonal part, \(H_\mathrm{corr}\), is treated explicitly.

@@

The OpenFOAM implementation of the numerical model and grid generation utilities are available at \url{https://github.com/hertzsprung/AtmosFOAM/tree/shaw-weller-2015-mwr}.  The test cases are located at \url{https://github.com/hertzsprung/AtmosTests/tree/shaw-weller-2015-mwr}.

@@

By \(t = \SI{10000}{\second}\), the tracer suffers from distortion on the BTF grid using the centred linear scheme and some artefacts remain about the mountain peak.  The tracer has spread vertically due to increased numerical errors when the tracer is transported between layers.  Distortions are reduced by using the fourth order scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}c), but the computational mode is seen as a grid-scale oscillation that travels in the opposite direction to the wind (figure~\ref{fig:advection-tracer}e).  The results from the centred linear and fourth order schemes are slightly worse than the respective results from \citet{schaer2002} (their figure 6a and figure 8) because of the difference in the discretisation of $\Psi$.

@@

The results of the horizontal advection test show that numerical errors are due either to misalignment of the flow with the grid, or to grid distortions.  Using the upwind-biased cubic scheme, distortions in the grid do not significantly distort the tracer.

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-advection-tracer/fig-advection-tracer.pdf}
%
	\caption{Horizontally advected tracer contours at \(t = \SI{0}{\second}\), \SI{5000}{\second} and \SI{10000}{\second} using (a) centred linear scheme on a BTF grid, the upwind-biased cubic scheme on (b) the cut cell grid, (d) the BTF grid, and (c) the fourth order scheme from \citet{schaer2002} on the BTF grid, with contour intervals every 0.1.  Errors on the BTF grid at \(t = \SI{10000}{\second}\) are shown for (e) the fourth order scheme from \citet{schaer2002}, and (f) the upwind-biased cubic scheme, with contour intervals every 0.01.  Negative contours denoted by dotted lines.  The terrain profile is also shown immediately above the $x$ axis.  Subfigures (c) and (e) produced using the modified version of the code from \citet{schaer2002}.}
	\label{fig:advection-tracer}
\end{figure*}

\subsection{Terrain following advection}
In the horizontal advection test, results were more accurate where the flow was aligned with the grid layers, and distortions in the BTF grid led to increased errors.  To determine the primary source of numerical error, we formulate a new tracer advection test in which the velocity field is everywhere tangential to the basic terrain following coordinate surfaces.  In this new test, flow is aligned with the BTF grid layers.  Misalignment on the SLEVE and cut cell grids is designed to challenge the advection scheme.

@@

\begin{table*}[t]
	\caption{Minimum and maximum tracer magnitudes and \(\ell^2\) error norms at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.  For the horizontal advection test, minimum and maximum values are given for the fourth order scheme using the modified code from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{l l l l S S S S}
\hline\hline
                  &                     &                  & Analytic & {BTF}   & {SLEVE}         & {Cut cell}     & {No terrain} \\
\hline
Horizontal        & Centred linear      & \(\ell^2\) error & 0        & 0.0210  & 0.00233         & 0.00224        & 0.00223     \\
                  &                     & min              & 0        & -0.275  & -0.0252         & -0.0251        & -0.0251     \\
                  &                     & max              & 1        & 0.925   & 0.985           & 0.985          & 0.985       \\
                  & Fourth order        & min              & 0        & -0.0926 & -0.00174        & {---}          & -0.00178    \\
                  &                     & max              & 1        & 1.00    & 0.984           & {---}          & 0.983       \\
                  & Cubic upwind-biased & \(\ell^2\) error & 0        & 0.00791 & 0.00108         & 0.000577       & 0.000576    \\
                  &                     & min              & 0        & -0.0446 & -0.0106         & -0.000674      & -0.00674    \\
                  &                     & max              & 1        & 0.925   & 0.982           & 0.983          & 0.983       \\
\hline
Terrain following & Centred linear      & \(\ell^2\) error & 0        & 0.00251 & 0.0173          & 13.6           & {---}        \\
		  &                     & min              & 0        & -0.0245 & -0.120          & -623           & {---}        \\
                  &                     & max              & 1        & 0.985   & 0.950           & 3480           & {---}        \\
		  & Cubic upwind-biased & \(\ell^2\) error & 0        & 0.00154 & 0.0120          & 0.0134         & {---}        \\
                  &                     & min              & 0        & -0.0110 & -0.0263         & -0.028         & {---}        \\
                  &                     & max              & 1        & 0.983   & 0.865           & 0.851          & {---}        \\
\hline
\end{tabular}
\end{table*}

@@

The test was integrated forward by 5 hours on the BTF, SLEVE and cut cell grids, and a regular grid with flat terrain.  Maximum vertical velocities are presented in figure~\ref{fig:resting}.  In agreement with \citet{klemp2011}, vertical velocities are larger on more distorted grids.  However, magnitudes are smaller comparing results on the terrain following grids with those from \citet{klemp2011}.  Using the model from \citet{weller-shahrokhi2014}, which includes a curl-free pressure gradient formulation, $w$ reaches a maximum of \SI{0.33}{\meter\per\second} on the BTF grid, compared with a maximum of \(\sim \SI{7}{\meter\per\second}\) found by \citet{klemp2011} using their improved horizontal pressure gradient formulation.

Unlike the result from \citet{klemp2011}, the SLEVE grid does not significantly reduce vertical velocities compared to the BTF grid.  However, errors are two orders of magnitude smaller on the cut cell grid with vertical velocities of \(\sim \SI{1e-4}{\meter\per\second}\).  The smallest error of \(\sim \SI{1e-10}{\meter\per\second}\) is found on the regular grid.

@@

Test results are compared between the BTF, SLEVE and cut cell grids.  Vertical velocities on the BTF, SLEVE and cut cell grids are visually indistinguishable (not shown).  They agree with the high resolution mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.

@@
